% ============================================================
% STAT 8289 Course Project - JASA Format
% Title: Performance and Interpretability Trade-offs in
%        Reinforcement Learning for Sepsis Treatment:
%        Comparing Offline and Online Approaches
% Authors: Zhiyu Cheng, Yalun Ding, Chuanhui Peng (Equal contribution)
% Date: October 2025
% ============================================================

\documentclass[11pt]{article}

% ============================================================
% PACKAGES
% ============================================================
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp}
\else
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}

% Formatting
\usepackage{ifthen}
\IfFileExists{microtype.sty}{\usepackage[]{microtype}}{}
\usepackage{xcolor}
\setlength{\emergencystretch}{3em}
\setcounter{secnumdepth}{5}

% Tables and figures
\usepackage{longtable,booktabs,array}
\usepackage{calc}
\usepackage{graphicx}
\usepackage{subcaption}

% Page layout (JASA format)
\addtolength{\oddsidemargin}{-.5in}
\addtolength{\evensidemargin}{-.1in}
\addtolength{\textwidth}{1in}
\addtolength{\textheight}{1.7in}
\addtolength{\topmargin}{-1in}

% Graphics
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\makeatletter
\def\fps@figure{htbp}
\makeatother

% Bibliography
\usepackage[]{natbib}
\bibliographystyle{agsm}

% Hyperlinks
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{}
\urlstyle{same}
\hypersetup{
  pdftitle={Performance and Interpretability Trade-offs in Reinforcement Learning for Sepsis Treatment: Comparing Offline and Online Approaches},
  pdfauthor={Zhiyu Cheng, Yalun Ding, Chuanhui Peng},
  pdfkeywords={Reinforcement Learning, Sepsis Treatment, Interpretability, Conservative Q-Learning, LEG Analysis, Offline RL, MIMIC-III},
  colorlinks=true,
  linkcolor={blue},
  filecolor={red},
  citecolor={blue},
  urlcolor={blue},
  pdfcreator={LaTeX via pandoc}
}

% Custom commands
\newcommand{\anon}{1}  % Set to 0 for anonymous version, 1 for author version

% ============================================================
% DOCUMENT START
% ============================================================
\begin{document}

\def\spacingset#1{\renewcommand{\baselinestretch}{#1}\small\normalsize}
\spacingset{1}

% ============================================================
% TITLE AND AUTHORS
% ============================================================
\if1\anon
{
  \title{\bf Performance and Interpretability Trade-offs in Reinforcement Learning for Sepsis Treatment: Comparing Offline and Online Approaches}
  \author{
    Zhiyu Cheng\thanks{Equal contribution. Authors listed alphabetically. The authors gratefully acknowledge the support of STAT 8289 - Reinforcement Learning course at George Washington University. This work uses the MIMIC-III database (Johnson et al., 2016) and builds upon the gym-sepsis environment (Raghu et al., 2017).}$^{1}$,
    Yalun Ding$^{1,*}$,
    Chuanhui Peng$^{1}$ \\
    $^{1}$Department of Statistics, George Washington University \\
    $^{*}$Corresponding author: yding@gwu.edu
  }
  \maketitle
} \fi

\if0\anon
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf Performance and Interpretability Trade-offs in Reinforcement Learning for Sepsis Treatment: Comparing Offline and Online Approaches}
\end{center}
  \medskip
} \fi

% ============================================================
% ABSTRACT
% ============================================================
\bigskip
\begin{abstract}
Sepsis remains a leading cause of mortality in critical care, and reinforcement learning (RL) offers a promising route to data-driven treatment policies. Yet clinical adoption is impeded by the prevailing assumption that interpretability inevitably compromises performance, and that online RL methods necessarily outperform offline approaches. We interrogate these trade-offs by comparing three offline RL methods (Behavior Cloning, Conservative Q-Learning, and Deep Q-Network trained on static datasets) with three online RL methods (Double DQN with Attention, Double DQN with Residual connections, and Soft Actor-Critic with environment interaction) using the gym-sepsis simulator—a MIMIC-III–derived environment for sepsis treatment. Policy quality is assessed through patient-survival rates stratified by Sequential Organ Failure Assessment (SOFA) states, while interpretability is quantified with Linearly Estimated Gradients (LEG), a model-agnostic feature-importance method.

Across 500 evaluation episodes, online RL achieves marginally higher overall survival (95.4\% for DDQN-Attention vs. 94.2\% for BC), with a 1.9 percentage point advantage on high-severity patients (90.5\% vs. 88.6\%). However, this modest performance gain comes at the cost of requiring extensive environment interaction during training—infeasible in clinical settings where patient safety is paramount. Interpretability analysis reveals that offline methods, particularly CQL, produce LEG saliency peaks of 40.06—roughly 600-fold larger than DQN's 0.069—highlighting clinically coherent emphasis on blood pressure and lactate levels. These results demonstrate that high-performing offline RL policies can deliver comparable survival rates with transparent decision logic and no patient risk during training, challenging the presumed necessity of online learning. Selecting offline algorithms such as CQL therefore offers a viable path toward regulatory-grade, clinician-trustworthy AI for sepsis management.
\end{abstract}

\noindent%
{\it Keywords:} Reinforcement Learning, Sepsis Treatment, Interpretability, Conservative Q-Learning, LEG Analysis, Offline RL, MIMIC-III

\vfill
\newpage
\spacingset{1} % Single line spacing for 25-page limit

% ============================================================
% 1. INTRODUCTION
% ============================================================
\input{sections/01_introduction.tex}


% ============================================================
% 2. RELATED WORK
% ============================================================
\input{sections/02_related.tex}


% ============================================================
% 3. PROBLEM FORMULATION
% ============================================================
\input{sections/03_problem.tex}


% ============================================================
% 4. METHODS
% ============================================================
\input{sections/04_methods.tex}


% ============================================================
% 5. RESULTS
% ============================================================
\input{sections/05_results.tex}


% ============================================================
% 6. DISCUSSION
% ============================================================
\input{sections/06_discussion.tex}


% ============================================================
% 7. CONCLUSION
% ============================================================
\input{sections/07_conclusion.tex}


% ============================================================
% AUTHOR CONTRIBUTIONS
% ============================================================
\section{Author Contributions}
\input{sections/08_contributions.tex}


% ============================================================
% DISCLOSURE & DATA AVAILABILITY
% ============================================================
\section{Disclosure Statement}

The authors declare no conflicts of interest.

\section{Data Availability Statement}

This study uses the MIMIC-III database \citep{johnson2016mimic3} and the gym-sepsis simulation environment (\url{https://github.com/gefeilin/gym-sepsis/tree/main/gym_sepsis/envs}). Code for replication is available at \url{https://github.com/akiani/gym-sepsis}.

% ============================================================
% APPENDIX
% ============================================================
\input{sections/09_appendix.tex}

% ============================================================
% REFERENCES
% ============================================================
\bibliography{references.bib}

\end{document}
